---
title: "Feature_Processing"
author: "Jannik Franken"
date: "18 5 2026"
output: html_document
---


```{r}
library(dplyr)
library(tidyr)
library(ggplot2)
library(purrr)
library(scales)
library(stats)
library(magrittr)
library(readr)
library(stringr)

library(MetaProViz)
```

```{r}

if (!dir.exists("../Results/Feature_processing")) {
  dir.create("../Results/Feature_processing", recursive = TRUE)
}
if (!dir.exists("../Results/Feature_processing/original")) {
  dir.create("../Results/Feature_processing/original", recursive = TRUE)
}
if (!dir.exists("../Results/Feature_processing/ID_QC")) {
  dir.create("../Results/Feature_processing/ID_QC", recursive = TRUE)
}
if (!dir.exists("../Results/Feature_processing/cleaned")) {
  dir.create("../Results/Feature_processing/cleaned", recursive = TRUE)
}
if (!dir.exists("../Results/Feature_processing/traverse")) {
  dir.create("../Results/Feature_processing/traverse", recursive = TRUE)
}
if (!dir.exists("../Results/Feature_processing/addEquivalent")) {
  dir.create("../Results/Feature_processing/addEquivalent", recursive = TRUE)
}
if (!dir.exists("../Results/Feature_processing/id_summary")) {
  dir.create("../Results/Feature_processing/id_summary", recursive = TRUE)
}
if (!dir.exists("../Results/Feature_processing/id_summary_plots")) {
  dir.create("../Results/Feature_processing/id_summary_plots", recursive = TRUE)
}
```


```{r}
########################
### HELPER functions for id_summary table and deltas

## Helper: extract one summary row from count_id() output
extract_id_counts <- function(count_obj, row_name) {
  ## count_obj must be an object returned by MetaProViz::count_id()
  
  tbl <- count_obj$Table
  
  ## Count how many features fall into each ID class
  counts_wide <- tbl %>%
    count(id_label, name = "n") %>%
    mutate(
      id_label = recode(
        id_label,
        "No ID" = "no_ID",
        "Single ID" = "Single_ID",
        "Multiple IDs" = "Multiple_IDs"
      )
    ) %>%
    pivot_wider(
      names_from = id_label,
      values_from = n,
      values_fill = 0
    )
  
  ## Make sure all expected columns exist
  ## This avoids errors if one class is absent in a given dataset
  for (col in c("no_ID", "Single_ID", "Multiple_IDs")) {
    if (!col %in% names(counts_wide)) {
      counts_wide[[col]] <- 0L
    }
  }
  
  ## Build one clean summary row
  tibble(
    row_name = row_name,
    no_ID = as.integer(counts_wide$no_ID[1]),
    Single_ID = as.integer(counts_wide$Single_ID[1]),
    Multiple_IDs = as.integer(counts_wide$Multiple_IDs[1]),
    Total_IDs = as.integer(sum(tbl$entry_count, na.rm = TRUE))
  )
}



## Helper: append one summary row to id_count_df
append_id_counts <- function(df, count_obj, row_name) {
  bind_rows(df, extract_id_counts(count_obj, row_name))
}

## Helper: calculate deltas within each database only

## Example:
## HMDB_cleaned - HMDB_Original
## KEGG_cleaned - KEGG_Original
## PUBCHEM_cleaned - PUBCHEM_Original

## This does not compare across databases.
calculate_id_deltas <- function(df, from_category, to_category) {
  
  df %>%
    separate(row_name, into = c("id_type", "category"), sep = "_", remove = FALSE) %>%
    select(-row_name) %>%
    pivot_wider(
      names_from = category,
      values_from = c(no_ID, Single_ID, Multiple_IDs, Total_IDs)
    ) %>%
    transmute(
      row_name = paste0("delta_", id_type, "_", from_category, "_", id_type, "_", to_category),
      no_ID = .data[[paste0("no_ID_", to_category)]] - .data[[paste0("no_ID_", from_category)]],
      Single_ID = .data[[paste0("Single_ID_", to_category)]] - .data[[paste0("Single_ID_", from_category)]],
      Multiple_IDs = .data[[paste0("Multiple_IDs_", to_category)]] - .data[[paste0("Multiple_IDs_", from_category)]],
      Total_IDs = .data[[paste0("Total_IDs_", to_category)]] - .data[[paste0("Total_IDs_", from_category)]]
    )
}

## Helper: count features with no HMDB, KEGG, or PUBCHEM
## This works on the original category dataframe, not on count_id() output.
## Example inputs:  MetaboliteIDs    or   MetaboliteID_Translated
count_none_all_ids <- function(df) {
  sum(
    (is.na(df$HMDB)    | trimws(as.character(df$HMDB)) %in% c("", "NA")) &
    (is.na(df$KEGG)    | trimws(as.character(df$KEGG)) %in% c("", "NA")) &
    (is.na(df$PUBCHEM) | trimws(as.character(df$PUBCHEM)) %in% c("", "NA")),
    na.rm = TRUE
  )
}

## Helper: append one row to none_summary_df
append_none_summary <- function(df, input_df, row_name) {
  bind_rows(
    df,
    tibble(
      row_name = row_name,
      No_database_ID = as.integer(count_none_all_ids(input_df))
    )
  )
}

## Helper: calculate delta for the "none at all" summary
## Example:  ## cleaned - Original
calculate_none_delta <- function(df, from_category, to_category) {
  
  tibble(
    row_name = paste0("delta_", from_category, "_", to_category),
    No_database_ID =
      df$No_database_ID[df$row_name == to_category] -
      df$No_database_ID[df$row_name == from_category]
  )
}

```

```{r}
load("../data/Shorthouse_2022.RData")
```


# Plan

Perform this logic for all 4 experiments without sub-versions
We will be working on the ...$features dataframes only.
Generally, assess ID state, then do something (check, traverse, addEquivalent),
re-assess ID state again

1. Assess ID state "original"

2. Perform ID QC seed_id_compatibility_check
3. Assess ID state "cleaned"

4. Perform MetaProViz::traverse_ids()
5. Assess ID state "traverse"

6. Perform MetaProViz::equivalent_id()
7. Assess ID state "addEquivalent"


```{r}
# 1. Assess ID state "original"

features_Shorthouse_original <- Shorthouse_2022$features %>%
    mutate(
        chebi = NA,
        pubchem = NA
    ) %>%
    mutate(
        across(any_of(c("hmdb", "kegg", "chebi", "pubchem")), ~ as.character(.x))
    )

# save original as csv
write.csv(
    features_Shorthouse_original,
    "../Results/Feature_processing/original/features_Shorthouse_original.csv",
    row.names = TRUE,
    fileEncoding = "UTF-8"
)

# assess original ID status
# using count_id and compare_pk

compare_pk_Shorthouse_original <-
    MetaProViz::compare_pk(
        data = list(Biocft = features_Shorthouse_original %>% dplyr::rename("Class" = "metabolite_class")),
        name_col = "name",
        metadata_info = list(Biocft = c("hmdb", "kegg", "chebi", "pubchem")),
        plot_name = "Shorthouse_2022 original - overlap of ID types",
        path = "../Results/Feature_processing/original"
    )
ggplot2::ggsave(
  filename = "../Results/Feature_processing/original/compare_pk_Shorthouse_original.svg",
  plot = compare_pk_Shorthouse_original$upset_plot,
  width = 12,
  height = 8
)

count_id_Shorthouse_original_HMDB <- 
    MetaProViz::count_id(
        features_Shorthouse_original, 
        delimiter = ";",
        "hmdb",
        fill_colors = c(`No ID` = "#006384", `Single ID` = "#9F0162", `Multiple IDs` = "#BEBEBE"),
        path = "../Results/Feature_processing/original"
    )
ggplot2::ggsave(
  filename = "../Results/Feature_processing/original/count_id_Shorthouse_original_HMDB.svg",
  plot = count_id_Shorthouse_original_HMDB$Plot,
  width = 12,
  height = 8
)

count_id_Shorthouse_original_KEGG <- 
    MetaProViz::count_id(
        features_Shorthouse_original, 
        delimiter = ";",
        "kegg",
        fill_colors = c(`No ID` = "#006384", `Single ID` = "#9F0162", `Multiple IDs` = "#BEBEBE"),
        path = "../Results/Feature_processing/original"
    )
ggplot2::ggsave(
  filename = "../Results/Feature_processing/original/count_id_Shorthouse_original_KEGG.svg",
  plot = count_id_Shorthouse_original_KEGG$Plot,
  width = 12,
  height = 8
)

count_id_Shorthouse_original_CHEBI <- 
    MetaProViz::count_id(
        features_Shorthouse_original, 
        delimiter = ";",
        "chebi",
        fill_colors = c(`No ID` = "#006384", `Single ID` = "#9F0162", `Multiple IDs` = "#BEBEBE"),
        path = "../Results/Feature_processing/original"
    )
ggplot2::ggsave(
  filename = "../Results/Feature_processing/original/count_id_Shorthouse_original_CHEBI.svg",
  plot = count_id_Shorthouse_original_CHEBI$Plot,
  width = 12,
  height = 8
)

count_id_Shorthouse_original_PUBCHEM <- 
    MetaProViz::count_id(
        features_Shorthouse_original, 
        delimiter = ";",
        "pubchem",
        fill_colors = c(`No ID` = "#006384", `Single ID` = "#9F0162", `Multiple IDs` = "#BEBEBE"),
        path = "../Results/Feature_processing/original"
    )
ggplot2::ggsave(
  filename = "../Results/Feature_processing/original/count_id_Shorthouse_original_PUBCHEM.svg",
  plot = count_id_Shorthouse_original_PUBCHEM$Plot,
  width = 12,
  height = 8
)
```



```{r}
# Initialise the main per-database summary dataframe
id_summary_Shorthouse <- tibble(
  row_name      = character(),
  no_ID         = integer(),
  Single_ID     = integer(),
  Multiple_IDs  = integer(),
  Total_IDs     = integer()
)

id_summary_Shorthouse <- append_id_counts(id_summary_Shorthouse, count_id_Shorthouse_original_HMDB,    "HMDB_original")
id_summary_Shorthouse <- append_id_counts(id_summary_Shorthouse, count_id_Shorthouse_original_KEGG,    "KEGG_original")
id_summary_Shorthouse <- append_id_counts(id_summary_Shorthouse, count_id_Shorthouse_original_CHEBI,   "CHEBI_original")
id_summary_Shorthouse <- append_id_counts(id_summary_Shorthouse, count_id_Shorthouse_original_KEGG,    "PUBCHEM_original")

none_summary_Shorthouse <- tibble(
  row_name = character(),
  No_database_ID = integer()
)

none_summary_Shorthouse <- append_none_summary(none_summary_Shorthouse, features_Shorthouse_original, "original")
```

```{r}
# 2. Perform ID QC seed_id_compatibility_check

# features_Shorthouse_compatibility_check <-
#     MetaProViz:::seed_id_compatibility_check(
#         data = features_Shorthouse_original,
#         id_types = c("HMDB", "KEGG", "CHEBI", "PUBCHEM"),
#         delimiter = ";",
#         verbose = TRUE
#     )
# save(features_Shorthouse_compatibility_check, file = "../data/features_Shorthouse_compatibility_check.RData")

load("../data/features_Shorthouse_compatibility_check.RData")

compatibility_by_feature_Shorthouse <- 
    features_Shorthouse_compatibility_check$ID_pair_compatibility %>%
    group_by(name) %>%
    summarise(
        n_pairs = n(),
        n_true = sum(pair_compatible == TRUE, na.rm = TRUE),
        n_false = sum(pair_compatible == FALSE, na.rm = TRUE),
        all_seed_ids_compatible = all_seed_ids_compatible,
        compatibility_class = case_when(
            all_seed_ids_compatible ~ "fully_compatible",
            !all_seed_ids_compatible & n_true > 0 & n_false > 0 ~ "partially_compatible",
            !all_seed_ids_compatible & n_true == 0 & n_false > 0 ~ "completely_incompatible",
            TRUE ~ NA_character_
        ),
        .groups = "drop"
    )

fully_compatible_Shorthouse <- compatibility_by_feature_Shorthouse %>% filter(compatibility_class == "fully_compatible")
partially_compatible_Shorthouse <- compatibility_by_feature_Shorthouse %>% filter(compatibility_class == "partially_compatible")
completely_incompatible_Shorthouse <- compatibility_by_feature_Shorthouse %>% filter(compatibility_class == "completely_incompatible")
# how many features are:
# fully_compatible?
length(unique(fully_compatible_Shorthouse$name)) # 495
# partially_compatible
length(unique(partially_compatible_Shorthouse$name)) # 209
# completely_incompatible
length(unique(completely_incompatible_Shorthouse$name)) # 392


# save ID QC results
write.csv(
    features_Shorthouse_compatibility_check$ID_pair_compatibility,
    "../Results/Feature_processing/ID_QC/features_Shorthouse_compatibility_check___ID_pair_compatibility.csv",
    row.names = TRUE,
    fileEncoding = "UTF-8"
)
write.csv(
    features_Shorthouse_compatibility_check$data_with_compatibility,
    "../Results/Feature_processing/ID_QC/features_Shorthouse_compatibility_check___data_with_compatibility.csv",
    row.names = TRUE,
    fileEncoding = "UTF-8"
)
write.csv(
    fully_compatible_Shorthouse,
    "../Results/Feature_processing/ID_QC/fully_compatible_Shorthouse.csv",
    row.names = TRUE,
    fileEncoding = "UTF-8"
)
write.csv(
    partially_compatible_Shorthouse,
    "../Results/Feature_processing/ID_QC/partially_compatible_Shorthouse.csv",
    row.names = TRUE,
    fileEncoding = "UTF-8"
)
write.csv(
    completely_incompatible_Shorthouse,
    "../Results/Feature_processing/ID_QC/completely_incompatible_Shorthouse.csv",
    row.names = TRUE,
    fileEncoding = "UTF-8"
)

# get ID_pair_compatibility rows for fully_compatible features
fully_compatible_metabolites <- unique(fully_compatible_Shorthouse$name)

fully_compatible_id_pair_rows <- 
    features_Shorthouse_compatibility_check$ID_pair_compatibility %>%
    filter(
        name %in% fully_compatible_metabolites
    )
# keep these as is to later merge together with the others


### handle partially_compatible IDs

# retrieve ID_pair_compatibility for partially_compatible features
partially_compatible_metabolites <- unique(partially_compatible_Shorthouse$name)

partially_compatible_id_pair_rows <- 
    features_Shorthouse_compatibility_check$ID_pair_compatibility %>%
    filter(
        name %in% partially_compatible_metabolites
    )

# keep only the matching id pairs through direct and secondary match, discard no_match ID pairs completely
partially_compatible_id_pair_rows_filtered <- 
    partially_compatible_id_pair_rows %>%
    filter(
        compatibility_path != "no_match"   #only keep matches
    )
# sanity check: still all features there?
length(unique(partially_compatible_id_pair_rows_filtered$name))

write.csv(
    partially_compatible_id_pair_rows_filtered,
    "../Results/Feature_processing/ID_QC/partially_compatible_id_pair_rows_filtered_Shorthouse.csv",
    row.names = TRUE,
    fileEncoding = "UTF-8"
)


### handle completely_incompatible IDs
completely_incompatible_metabolites <- unique(completely_incompatible_Shorthouse$name)

completely_incompatible_metabolites_id_pair_rows <- 
    features_Shorthouse_compatibility_check$ID_pair_compatibility %>%
    filter(
        name %in% completely_incompatible_metabolites
    )

completely_incompatible_data_with_compatibility <- features_Shorthouse_compatibility_check$data_with_compatibility %>%
    filter(
        name %in% completely_incompatible_metabolites
    )

## these are too many to manually review. Therefore for now I am going to do this:
# - if there is one KEGG ID in addition to multiple HMDB IDs, keep only the KEGG ID
# - if there are only two HMDB IDs, keep both
# - if there are multiple HMDB and two KEGG IDs, keep first HMDB and first KEGG
# - if there are only many HMDB IDs, keep the first one
# - if there is only one HMDB ID, and one KEGG ID, keep it and discard KEGG

split_ids <- function(x) {
    if (length(x) == 0) return(character(0))
    x <- as.character(x)
    x <- x[!is.na(x) & trimws(x) != "" & toupper(trimws(x)) != "NA"]
    if (length(x) == 0) return(character(0))
    vals <- unlist(strsplit(x[[1]], ";", fixed = TRUE))
    vals <- trimws(vals)
    vals <- vals[!is.na(vals) & vals != "" & toupper(vals) != "NA"]
    unique(vals)
}

first_non_empty <- function(x) {
    x <- x[!is.na(x) & trimws(as.character(x)) != "" & toupper(trimws(as.character(x))) != "NA"]
    if (length(x) == 0) return(NA_character_)
    as.character(x[[1]])
}

completely_incompatible_data_with_compatibility_processed <-
    completely_incompatible_data_with_compatibility %>%
    group_by(ionIdx, name) %>%
    mutate(
        hmdb_raw = first_non_empty(hmdb),
        kegg_raw = first_non_empty(kegg),
        hmdb_ids = list(split_ids(hmdb_raw)),
        kegg_ids = list(split_ids(kegg_raw)),
        n_hmdb = length(hmdb_ids[[1]]),
        n_kegg = length(kegg_ids[[1]]),
        hmdb_new = case_when(
            n_kegg == 1 & n_hmdb > 1 ~ NA_character_,
            n_kegg == 0 & n_hmdb == 2 ~ paste(hmdb_ids[[1]][1:2], collapse = ";"),
            n_kegg == 1 & n_hmdb == 1 ~ hmdb_ids[[1]][1],
            n_kegg == 2 & n_hmdb > 1 ~ hmdb_ids[[1]][1],
            n_kegg == 0 & n_hmdb > 2 ~ hmdb_ids[[1]][1],
            TRUE ~ hmdb_raw
        ),
        kegg_new = case_when(
            n_kegg == 1 & n_hmdb > 1 ~ kegg_ids[[1]][1],
            n_kegg == 2 & n_hmdb > 1 ~ kegg_ids[[1]][1],
            TRUE ~ kegg_raw
        ),
        hmdb = if_else(is.na(hmdb_new) | hmdb_new == "", NA_character_, hmdb_new),
        kegg = if_else(is.na(kegg_new) | kegg_new == "", NA_character_, kegg_new),
        HMDB = hmdb,
        KEGG = kegg
    ) %>%
    ungroup() %>%
    select(
        -hmdb_raw, -kegg_raw, -hmdb_ids, -kegg_ids, -n_hmdb, -n_kegg, -hmdb_new, -kegg_new, -HMDB, -KEGG, -CHEBI, -PUBCHEM
    ) %>% 
    mutate(
        sub_tbl = "completely_incompatible"
    )

write.csv(
    completely_incompatible_data_with_compatibility_processed,
    "../Results/Feature_processing/ID_QC/completely_incompatible_data_with_compatibility_processed_Shorthouse.csv",
    row.names = TRUE,
    fileEncoding = "UTF-8"
)

# merge back together ID_pair_compatibility for fully_compatible, partially_compatible
merge_completely_and_partially_compatible_ID_pair_rows <- 
    rbind(
        fully_compatible_id_pair_rows %>% mutate(sub_tbl = "fully_compatible"),
        partially_compatible_id_pair_rows_filtered %>% mutate(sub_tbl = "partially_compatible")
    )
length(unique(merge_completely_and_partially_compatible_ID_pair_rows$ionIdx)) # 704 is correct

# summarize by original row id, rbind to completely_incompatible_data_with_compatibility_processed and construct the features_Shorthouse_cleaned from this

# join back together based on "original_row_id"

features_Shorthouse_cleaned <- 
    rbind(
        merge_completely_and_partially_compatible_ID_pair_rows %>%
            select(-hmdb_kegg) %>%
            group_by(original_row_id) %>%
            summarise(
                name       = first(name),
                ionIdx     = first(ionIdx),
                ionMz      = first(ionMz),
                score      = first(score),
                formula    = first(formula),
                ion        = first(ion),
                
                hmdb = {
                    x <- trimws(as.character(HMDB))
                    x[x == ""] <- NA_character_
                    x <- unique(na.omit(x))
                    if (length(x) == 0) NA_character_ else paste(x, collapse = ", ")
                },
                kegg = {
                    x <- trimws(as.character(KEGG))
                    x[x == ""] <- NA_character_
                    x <- unique(na.omit(x))
                    if (length(x) == 0) NA_character_ else paste(x, collapse = ", ")
                },
                chebi = {
                    x <- trimws(as.character(CHEBI))
                    x[x == ""] <- NA_character_
                    x <- unique(na.omit(x))
                    if (length(x) == 0) NA_character_ else paste(x, collapse = ", ")
                },
                pubchem = {
                    x <- trimws(as.character(PUBCHEM))
                    x[x == ""] <- NA_character_
                    x <- unique(na.omit(x))
                    if (length(x) == 0) NA_character_ else paste(x, collapse = ", ")
                },
                
                sub_tbl = first(sub_tbl),
                metabolite_class = first(metabolite_class),
                
                .groups = "drop"
            ) %>%
            select(
                - original_row_id
            ),
        completely_incompatible_data_with_compatibility_processed %>%
            select(-all_seed_ids_compatible, -hmdb_kegg)
    ) %>%
    arrange(ionIdx)

# save as csv
write.csv(
    features_Shorthouse_cleaned,
    "../Results/Feature_processing/cleaned/features_Shorthouse_cleaned.csv",
    row.names = TRUE,
    fileEncoding = "UTF-8"
)

```

```{r}
# 3. Assess ID state "cleaned"

# assess cleaned ID status
# using count_id and compare_pk

compare_pk_Shorthouse_cleaned <-
    MetaProViz::compare_pk(
        data = list(Biocft = features_Shorthouse_cleaned %>% dplyr::rename("Class" = "metabolite_class")),
        name_col = "name",
        metadata_info = list(Biocft = c("hmdb", "kegg", "chebi", "pubchem")),
        plot_name = "Shorthouse_2022 cleaned - overlap of ID types",
        path = "../Results/Feature_processing/cleaned"
    )
ggplot2::ggsave(
  filename = "../Results/Feature_processing/cleaned/compare_pk_Shorthouse_cleaned.svg",
  plot = compare_pk_Shorthouse_cleaned$upset_plot,
  width = 12,
  height = 8
)

count_id_Shorthouse_cleaned_HMDB <- 
    MetaProViz::count_id(
        features_Shorthouse_cleaned, 
        delimiter = ";",
        "hmdb",
        fill_colors = c(`No ID` = "#006384", `Single ID` = "#9F0162", `Multiple IDs` = "#BEBEBE"),
        path = "../Results/Feature_processing/cleaned"
    )
ggplot2::ggsave(
  filename = "../Results/Feature_processing/cleaned/count_id_Shorthouse_cleaned_HMDB.svg",
  plot = count_id_Shorthouse_cleaned_HMDB$Plot,
  width = 12,
  height = 8
)

count_id_Shorthouse_cleaned_KEGG <- 
    MetaProViz::count_id(
        features_Shorthouse_cleaned, 
        delimiter = ";",
        "kegg",
        fill_colors = c(`No ID` = "#006384", `Single ID` = "#9F0162", `Multiple IDs` = "#BEBEBE"),
        path = "../Results/Feature_processing/cleaned"
    )
ggplot2::ggsave(
  filename = "../Results/Feature_processing/cleaned/count_id_Shorthouse_cleaned_KEGG.svg",
  plot = count_id_Shorthouse_cleaned_KEGG$Plot,
  width = 12,
  height = 8
)

count_id_Shorthouse_cleaned_CHEBI <- 
    MetaProViz::count_id(
        features_Shorthouse_cleaned, 
        delimiter = ";",
        "chebi",
        fill_colors = c(`No ID` = "#006384", `Single ID` = "#9F0162", `Multiple IDs` = "#BEBEBE"),
        path = "../Results/Feature_processing/cleaned"
    )
ggplot2::ggsave(
  filename = "../Results/Feature_processing/cleaned/count_id_Shorthouse_cleaned_CHEBI.svg",
  plot = count_id_Shorthouse_cleaned_CHEBI$Plot,
  width = 12,
  height = 8
)

count_id_Shorthouse_cleaned_PUBCHEM <- 
    MetaProViz::count_id(
        features_Shorthouse_cleaned, 
        delimiter = ";",
        "pubchem",
        fill_colors = c(`No ID` = "#006384", `Single ID` = "#9F0162", `Multiple IDs` = "#BEBEBE"),
        path = "../Results/Feature_processing/cleaned"
    )
ggplot2::ggsave(
  filename = "../Results/Feature_processing/cleaned/count_id_Shorthouse_cleaned_PUBCHEM.svg",
  plot = count_id_Shorthouse_cleaned_PUBCHEM$Plot,
  width = 12,
  height = 8
)

```

```{r}
# append the new "cleaned" ID counts

id_summary_Shorthouse <- append_id_counts(id_summary_Shorthouse, count_id_Shorthouse_cleaned_HMDB,    "HMDB_cleaned")
id_summary_Shorthouse <- append_id_counts(id_summary_Shorthouse, count_id_Shorthouse_cleaned_KEGG,    "KEGG_cleaned")
id_summary_Shorthouse <- append_id_counts(id_summary_Shorthouse, count_id_Shorthouse_cleaned_CHEBI,   "CHEBI_cleaned")
id_summary_Shorthouse <- append_id_counts(id_summary_Shorthouse, count_id_Shorthouse_cleaned_PUBCHEM, "PUBCHEM_cleaned")

## Calculate delta rows
delta_df_Shorthouse <- calculate_id_deltas(id_summary_Shorthouse, "original", "cleaned")

## cleaned category dataframe
none_summary_Shorthouse <- append_none_summary(none_summary_Shorthouse, features_Shorthouse_cleaned, "cleaned")

## Calculate delta row
none_delta_Shorthouse <- calculate_none_delta(none_summary_Shorthouse, "original", "cleaned")
```


```{r}
# 4. Perform MetaProViz::traverse_ids()

# features_Shorthouse_traverse_ids <-
#     MetaProViz::traverse_ids(
#         data = features_Shorthouse_cleaned,
#         id_types = c("HMDB", "KEGG", "CHEBI", "PUBCHEM"),
#         delimiter = ";",
#         path = "../Results/Feature_processing/traverse"
#     )
# 
# save(features_Shorthouse_traverse_ids, file = "../data/features_Shorthouse_traverse_ids.RData")

load("../data/features_Shorthouse_traverse_ids.RData")

traverse_Shorthouse_ExpandedDF <- features_Shorthouse_traverse_ids$ExpandedDF
traverse_Shorthouse_ID_pair_compatibility <- features_Shorthouse_traverse_ids$ID_pair_compatibility



# create new feature table

features_Shorthouse_traverse <- 
    merge(
        features_Shorthouse_cleaned %>%
            select(
                - hmdb,
                - kegg,
                - chebi,
                - pubchem
            ),
        traverse_Shorthouse_ExpandedDF %>%
            select(
                name,
                hmdb,
                kegg,
                chebi,
                pubchem,
                n_HMDB_translated,
                n_KEGG_translated,
                n_CHEBI_translated,
                n_PUBCHEM_translated,
                HMDB_translated,
                KEGG_translated,
                CHEBI_translated,
                PUBCHEM_translated
            ),
        by.x = "name",
        by.y = "name",
        all.x = TRUE
    ) %>%
    mutate(
        hmdb = case_when(
            n_HMDB_translated < 50 ~ HMDB_translated,
            TRUE ~ hmdb
        ),
        kegg = case_when(
            n_KEGG_translated < 50 ~ KEGG_translated,
            TRUE ~ kegg
        ),
        chebi = case_when(
            n_CHEBI_translated < 50 ~ CHEBI_translated,
            TRUE ~ chebi
        ),
        pubchem = case_when(
            n_PUBCHEM_translated < 50 ~ PUBCHEM_translated,
            TRUE ~ pubchem
        )
    ) %>%
    mutate(
        all_ids = pmap_chr(
            pick(hmdb, kegg, chebi, pubchem),
            ~ {
                c(...) %>%
                    .[!is.na(.)] %>%
                    stringr::str_split(";") %>%
                    unlist() %>%
                    stringr::str_trim() %>%
                    .[. != ""] %>%
                    unique() %>%
                    {\(vals) if (length(vals) == 0) NA_character_ else paste(vals, collapse = ";")}()
            }
        )
    ) %>%
    select(
        - HMDB_translated,
        - KEGG_translated,
        - CHEBI_translated,
        - PUBCHEM_translated,
        - n_HMDB_translated,
        - n_KEGG_translated,
        - n_CHEBI_translated,
        - n_PUBCHEM_translated
    )

# save as csv
write.csv(
    features_Shorthouse_traverse,
    "../Results/Feature_processing/traverse/features_Shorthouse_traverse.csv",
    row.names = TRUE,
    fileEncoding = "UTF-8"
)

```

```{r}
# 5. Assess ID state "traverse"

# assess original ID status
# using count_id and compare_pk

compare_pk_Shorthouse_traverse <-
    MetaProViz::compare_pk(
        data = list(Biocft = features_Shorthouse_traverse %>% dplyr::rename("Class" = "metabolite_class")),
        name_col = "name",
        metadata_info = list(Biocft = c("hmdb", "kegg", "chebi", "pubchem")),
        plot_name = "Shorthouse_2022 traverse - overlap of ID types",
        path = "../Results/Feature_processing/traverse"
    )
ggplot2::ggsave(
  filename = "../Results/Feature_processing/traverse/compare_pk_Shorthouse_traverse.svg",
  plot = compare_pk_Shorthouse_traverse$upset_plot,
  width = 12,
  height = 8
)

count_id_Shorthouse_traverse_HMDB <- 
    MetaProViz::count_id(
        features_Shorthouse_traverse, 
        delimiter = ";",
        "hmdb",
        fill_colors = c(`No ID` = "#006384", `Single ID` = "#9F0162", `Multiple IDs` = "#BEBEBE"),
        path = "../Results/Feature_processing/traverse"
    )
ggplot2::ggsave(
  filename = "../Results/Feature_processing/traverse/count_id_Shorthouse_traverse_HMDB.svg",
  plot = count_id_Shorthouse_traverse_HMDB$Plot,
  width = 12,
  height = 8
)

count_id_Shorthouse_traverse_KEGG <- 
    MetaProViz::count_id(
        features_Shorthouse_traverse, 
        delimiter = ";",
        "kegg",
        fill_colors = c(`No ID` = "#006384", `Single ID` = "#9F0162", `Multiple IDs` = "#BEBEBE"),
        path = "../Results/Feature_processing/traverse"
    )
ggplot2::ggsave(
  filename = "../Results/Feature_processing/traverse/count_id_Shorthouse_traverse_KEGG.svg",
  plot = count_id_Shorthouse_traverse_KEGG$Plot,
  width = 12,
  height = 8
)

count_id_Shorthouse_traverse_CHEBI <- 
    MetaProViz::count_id(
        features_Shorthouse_traverse, 
        delimiter = ";",
        "chebi",
        fill_colors = c(`No ID` = "#006384", `Single ID` = "#9F0162", `Multiple IDs` = "#BEBEBE"),
        path = "../Results/Feature_processing/traverse"
    )
ggplot2::ggsave(
  filename = "../Results/Feature_processing/traverse/count_id_Shorthouse_traverse_CHEBI.svg",
  plot = count_id_Shorthouse_traverse_CHEBI$Plot,
  width = 12,
  height = 8
)

count_id_Shorthouse_traverse_PUBCHEM <- 
    MetaProViz::count_id(
        features_Shorthouse_traverse, 
        delimiter = ";",
        "pubchem",
        fill_colors = c(`No ID` = "#006384", `Single ID` = "#9F0162", `Multiple IDs` = "#BEBEBE"),
        path = "../Results/Feature_processing/traverse"
    )
ggplot2::ggsave(
  filename = "../Results/Feature_processing/traverse/count_id_Shorthouse_traverse_PUBCHEM.svg",
  plot = count_id_Shorthouse_traverse_PUBCHEM$Plot,
  width = 12,
  height = 8
)

```

```{r}
# append the new "traverse" ID counts

id_summary_Shorthouse <- append_id_counts(id_summary_Shorthouse, count_id_Shorthouse_traverse_HMDB,    "HMDB_traverse")
id_summary_Shorthouse <- append_id_counts(id_summary_Shorthouse, count_id_Shorthouse_traverse_KEGG,    "KEGG_traverse")
id_summary_Shorthouse <- append_id_counts(id_summary_Shorthouse, count_id_Shorthouse_traverse_CHEBI,   "CHEBI_traverse")
id_summary_Shorthouse <- append_id_counts(id_summary_Shorthouse, count_id_Shorthouse_traverse_PUBCHEM, "PUBCHEM_traverse")

## Calculate delta rows
delta_df_Shorthouse <- rbind(
  delta_df_Shorthouse,
  calculate_id_deltas(id_summary_Shorthouse, "original", "traverse"),
  calculate_id_deltas(id_summary_Shorthouse, "cleaned", "traverse")
)

## cleaned category dataframe
none_summary_Shorthouse <- append_none_summary(none_summary_Shorthouse, features_Shorthouse_traverse, "traverse")

## Calculate delta row
none_delta_Shorthouse <- rbind(
  none_delta_Shorthouse,
  calculate_none_delta(none_summary_Shorthouse, "original", "traverse"),
  calculate_none_delta(none_summary_Shorthouse, "cleaned", "traverse")
)
```

```{r}
# 6. Perform MetaProViz::equivalent_id()

## HMDB

Input_HMDB_Shorthouse <- features_Shorthouse_traverse %>%
    select(
        name,
        hmdb
    ) %>%
    separate_rows(
        hmdb, sep = ";"
    ) %>%
    mutate(
        hmdb = str_trim(hmdb)
    ) %>%
    filter(
        hmdb != ""
    )

length(which(duplicated(Input_HMDB_Shorthouse$hmdb)))  # 11301
length(which(!duplicated(Input_HMDB_Shorthouse$hmdb)))  # 3185

Input_HMDB_Shorthouse_distinct <- Input_HMDB_Shorthouse %>% distinct(hmdb, .keep_all = TRUE)


equivalent_id_res_Shorthouse_HMDB <- 
    MetaProViz::equivalent_id(
        data = Input_HMDB_Shorthouse_distinct,
        metadata_info = c(InputID = "hmdb"),
        from = "hmdb",
        path = "../Results/Feature_processing/addEquivalent"
    )

equivalent_id_res_Shorthouse_HMDB_clean <-
    equivalent_id_res_Shorthouse_HMDB %>%
    group_by(hmdb) %>%
    summarise(
        across(
            c(PotentialAdditionalIDs, AllIDs), ~{
                ids <- strsplit(as.character(.x), ",\\s*")
                ids <- trimws(unlist(ids, use.names = FALSE))
                ids <- ids[!is.na(ids) & ids != "" & ids != "NA"]
                if (length(ids) == 0) NA_character_ else paste(unique(ids), collapse = ";")
            }
        ),
        .groups = "drop"
    )

# now map the cleaned output from equivalent_id back to the metabolite
output_HMDB_Shorthouse <- 
    Input_HMDB_Shorthouse %>%
    left_join(
        equivalent_id_res_Shorthouse_HMDB_clean %>%
            select(
                hmdb,
                PotentialAdditionalIDs,
                AllIDs
            ),
        by = "hmdb"
      ) %>%
      group_by(name) %>%
      summarise(
          hmdb = {
              ids <- unlist(strsplit(paste(AllIDs, collapse = ";"), ";\\s*"))
              ids <- trimws(ids)
              ids <- unique(ids[!is.na(ids) & ids != "" & ids != "NA"])
              if (length(ids) == 0) NA_character_ else paste(ids, collapse = ";")
          },
          .groups = "drop"
      ) %>%
    mutate(
        n_expansion = lengths(strsplit(hmdb, ";"))
    )



## KEGG

Input_KEGG_Shorthouse <- features_Shorthouse_traverse %>%
    select(
        name,
        kegg
    ) %>%
    separate_rows(
        kegg, sep = ";"
    ) %>%
    mutate(
        kegg = str_trim(kegg)
    ) %>%
    filter(
        kegg != ""
    )

length(which(duplicated(Input_KEGG_Shorthouse$kegg)))  # 193
length(which(!duplicated(Input_KEGG_Shorthouse$kegg)))  # 890

Input_KEGG_Shorthouse_distinct <- Input_KEGG_Shorthouse %>% distinct(kegg, .keep_all = TRUE)


equivalent_id_res_Shorthouse_KEGG <- 
    MetaProViz::equivalent_id(
        data = Input_KEGG_Shorthouse_distinct,
        metadata_info = c(InputID = "kegg"),
        from = "kegg",
        path = "../Results/Feature_processing/addEquivalent"
    )

equivalent_id_res_Shorthouse_KEGG_clean <-
    equivalent_id_res_Shorthouse_KEGG %>%
    group_by(kegg) %>%
    summarise(
        across(
            c(PotentialAdditionalIDs, AllIDs), ~{
                ids <- strsplit(as.character(.x), ",\\s*")
                ids <- trimws(unlist(ids, use.names = FALSE))
                ids <- ids[!is.na(ids) & ids != "" & ids != "NA"]
                if (length(ids) == 0) NA_character_ else paste(unique(ids), collapse = ";")
            }
        ),
        .groups = "drop"
    )

# now map the cleaned output from equivalent_id back to the metabolite
output_KEGG_Shorthouse <- 
    Input_KEGG_Shorthouse %>%
    left_join(
        equivalent_id_res_Shorthouse_KEGG_clean %>%
            select(
                kegg,
                PotentialAdditionalIDs,
                AllIDs
            ),
        by = "kegg"
      ) %>%
      group_by(name) %>%
      summarise(
          kegg = {
              ids <- unlist(strsplit(paste(AllIDs, collapse = ";"), ";\\s*"))
              ids <- trimws(ids)
              ids <- unique(ids[!is.na(ids) & ids != "" & ids != "NA"])
              if (length(ids) == 0) NA_character_ else paste(ids, collapse = ";")
          },
          .groups = "drop"
      ) %>%
    mutate(
        n_expansion = lengths(strsplit(kegg, ";"))
    )



## CHEBI

Input_CHEBI_Shorthouse <- features_Shorthouse_traverse %>%
    select(
        name,
        chebi
    ) %>%
    separate_rows(
        chebi, sep = ";"
    ) %>%
    mutate(
        chebi = str_trim(chebi)
    ) %>%
    filter(
        chebi != ""
    ) %>%
    mutate(   # remove "CHEBI:" prefix for equivalent_id() function
         chebi = str_remove(str_to_upper(chebi), "^CHEBI:")
    )

length(which(duplicated(Input_CHEBI_Shorthouse$chebi)))  # 4744
length(which(!duplicated(Input_CHEBI_Shorthouse$chebi)))  # 2578

Input_CHEBI_Shorthouse_distinct <- Input_CHEBI_Shorthouse %>% distinct(chebi, .keep_all = FALSE)


equivalent_id_res_Shorthouse_CHEBI <- 
    MetaProViz::equivalent_id(
        data = Input_CHEBI_Shorthouse_distinct,
        metadata_info = c(InputID = "chebi"),
        from = "chebi",
        path = "../Results/Feature_processing/addEquivalent"
    )

equivalent_id_res_Shorthouse_CHEBI_clean <-
    equivalent_id_res_Shorthouse_CHEBI %>%
    group_by(chebi) %>%
    summarise(
        across(
            c(PotentialAdditionalIDs, AllIDs), ~{
                ids <- strsplit(as.character(.x), ",\\s*")
                ids <- trimws(unlist(ids, use.names = FALSE))
                ids <- ids[!is.na(ids) & ids != "" & ids != "NA"]
                if (length(ids) == 0) NA_character_ else paste(unique(ids), collapse = ";")
            }
        ),
        .groups = "drop"
    )

# now map the cleaned output from equivalent_id back to the metabolite
output_CHEBI_Shorthouse <- 
    Input_CHEBI_Shorthouse %>%
    left_join(
        equivalent_id_res_Shorthouse_CHEBI_clean %>%
            select(
                chebi,
                PotentialAdditionalIDs,
                AllIDs
            ),
        by = "chebi"
    ) %>%
    group_by(name) %>%
    summarise(
        chebi = {
            ids <- unlist(strsplit(paste(AllIDs, collapse = ";"), ";\\s*"))
            ids <- trimws(ids)
            ids <- unique(ids[!is.na(ids) & ids != "" & ids != "NA"])
            if (length(ids) == 0) NA_character_ else paste(ids, collapse = ";")
        },
        .groups = "drop"
    ) %>%
    mutate(  # get CHEBI: prefix back
        chebi = if_else(
            is.na(chebi) | chebi == "",
            chebi,
            purrr::map_chr(str_split(chebi, "\\s*;\\s*"), ~ {
                ids <- str_trim(.x)
                ids <- ids[ids != ""]
                ids <- if_else(str_detect(ids, "^CHEBI:"), ids, paste0("CHEBI:", ids))
                paste(ids, collapse = ";")
            })
        )
    ) %>%
    mutate(
        n_expansion = lengths(strsplit(chebi, ";"))
    )



## PUBCHEM

Input_PUBCHEM_Shorthouse <- features_Shorthouse_traverse %>%
    select(
        name,
        pubchem
    ) %>%
    separate_rows(
        pubchem, sep = ";"
    ) %>%
    mutate(
        pubchem = str_trim(pubchem)
    ) %>%
    filter(
        pubchem != ""
    ) %>%
    mutate(   # remove "CID:" prefix for equivalent_id() function
         pubchem = str_remove(str_to_upper(pubchem), "^CID")
    )

length(which(duplicated(Input_PUBCHEM_Shorthouse$pubchem)))  # 10594
length(which(!duplicated(Input_PUBCHEM_Shorthouse$pubchem)))  # 3060

Input_PUBCHEM_Shorthouse_distinct <- Input_PUBCHEM_Shorthouse %>% distinct(pubchem, .keep_all = FALSE)


equivalent_id_res_Shorthouse_PUBCHEM <- 
    MetaProViz::equivalent_id(
        data = Input_PUBCHEM_Shorthouse_distinct,
        metadata_info = c(InputID = "pubchem"),
        from = "pubchem",
        path = "../Results/Feature_processing/addEquivalent"
    )

equivalent_id_res_Shorthouse_PUBCHEM_clean <-
    equivalent_id_res_Shorthouse_PUBCHEM %>%
    group_by(pubchem) %>%
    summarise(
        across(
            c(PotentialAdditionalIDs, AllIDs), ~{
                ids <- strsplit(as.character(.x), ",\\s*")
                ids <- trimws(unlist(ids, use.names = FALSE))
                ids <- ids[!is.na(ids) & ids != "" & ids != "NA"]
                if (length(ids) == 0) NA_character_ else paste(unique(ids), collapse = ";")
            }
        ),
        .groups = "drop"
    )

# now map the cleaned output from equivalent_id back to the metabolite
output_PUBCHEM_Shorthouse <- 
    Input_PUBCHEM_Shorthouse %>%
    left_join(
        equivalent_id_res_Shorthouse_PUBCHEM_clean %>%
            select(
                pubchem,
                PotentialAdditionalIDs,
                AllIDs
            ),
        by = "pubchem"
    ) %>%
    group_by(name) %>%
    summarise(
        pubchem = {
            ids <- unlist(strsplit(paste(AllIDs, collapse = ";"), ";\\s*"))
            ids <- trimws(ids)
            ids <- unique(ids[!is.na(ids) & ids != "" & ids != "NA"])
            if (length(ids) == 0) NA_character_ else paste(ids, collapse = ";")
        },
        .groups = "drop"
    ) %>%
    mutate(  # get CHEBI: prefix back
        pubchem = if_else(
            is.na(pubchem) | pubchem == "",
            pubchem,
            purrr::map_chr(str_split(pubchem, "\\s*;\\s*"), ~ {
                ids <- str_trim(.x)
                ids <- ids[ids != ""]
                ids <- if_else(str_detect(ids, "^CID"), ids, paste0("CID", ids))
                paste(ids, collapse = ";")
            })
        )
    ) %>%
    mutate(
        n_expansion = lengths(strsplit(pubchem, ";"))
    )


```

```{r}

features_Shorthouse_addEquivalent <-
    features_Shorthouse_traverse %>%
    left_join(
        output_HMDB_Shorthouse %>% 
            filter(n_expansion < 50) %>%
            select(name, hmdb_new = hmdb),
        by = "name"
    ) %>%
    left_join(
        output_KEGG_Shorthouse %>% 
            filter(n_expansion < 50) %>%
            select(name, kegg_new = kegg),
        by = "name"
    ) %>%
    left_join(
        output_CHEBI_Shorthouse %>% 
            filter(n_expansion < 50) %>%
            select(name, chebi_new = chebi),
        by = "name"
    ) %>%
    left_join(
        output_PUBCHEM_Shorthouse %>% 
            filter(n_expansion < 50) %>%
            select(name, pubchem_new = pubchem),
        by = "name"
    ) %>%
    mutate(
        hmdb    = coalesce(hmdb_new, hmdb),
        kegg    = coalesce(kegg_new, kegg),
        chebi   = coalesce(chebi_new, chebi),
        pubchem = coalesce(pubchem_new, pubchem)
    ) %>%
    select(-hmdb_new, -kegg_new, -chebi_new, -pubchem_new) %>%
    relocate(
        hmdb, kegg, chebi, pubchem, .after = name
    ) %>%
    mutate(
        all_ids = pmap_chr( 
            pick(hmdb, kegg, chebi, pubchem),
            ~ {
                c(...) %>%
                    .[!is.na(.)] %>%
                    stringr::str_split(";") %>%
                    unlist() %>%
                    stringr::str_trim() %>%
                    .[. != ""] %>%
                    unique() %>%
                    {\(vals) if (length(vals) == 0) NA_character_ else paste(vals, collapse = ";")}()
            }
        )
    )

```

```{r}
# save as csv
write.csv(
    features_Shorthouse_addEquivalent,
    "../Results/Feature_processing/addEquivalent/features_Shorthouse_addEquivalent.csv",
    row.names = TRUE,
    fileEncoding = "UTF-8"
)
```

```{r}
# 7. Assess ID state "addEquivalent"


compare_pk_Shorthouse_addEquivalent <-
    MetaProViz::compare_pk(
        data = list(Biocft = features_Shorthouse_addEquivalent %>% dplyr::rename("Class" = "metabolite_class")),
        name_col = "name",
        metadata_info = list(Biocft = c("hmdb", "kegg", "chebi", "pubchem")),
        plot_name = "Shorthouse_2022 addEquivalent - overlap of ID types",
        path = "../Results/Feature_processing/addEquivalent"
    )
ggplot2::ggsave(
  filename = "../Results/Feature_processing/addEquivalent/compare_pk_Shorthouse_addEquivalent.svg",
  plot = compare_pk_Shorthouse_traverse$upset_plot,
  width = 12,
  height = 8
)

count_id_Shorthouse_addEquivalent_HMDB <- 
    MetaProViz::count_id(
        features_Shorthouse_addEquivalent, 
        delimiter = ";",
        "hmdb",
        fill_colors = c(`No ID` = "#006384", `Single ID` = "#9F0162", `Multiple IDs` = "#BEBEBE"),
        path = "../Results/Feature_processing/addEquivalent"
    )
ggplot2::ggsave(
  filename = "../Results/Feature_processing/addEquivalent/count_id_Shorthouse_addEquivalent_HMDB.svg",
  plot = count_id_Shorthouse_addEquivalent_HMDB$Plot,
  width = 12,
  height = 8
)

count_id_Shorthouse_addEquivalent_KEGG <- 
    MetaProViz::count_id(
        features_Shorthouse_addEquivalent, 
        delimiter = ";",
        "kegg",
        fill_colors = c(`No ID` = "#006384", `Single ID` = "#9F0162", `Multiple IDs` = "#BEBEBE"),
        path = "../Results/Feature_processing/addEquivalent"
    )
ggplot2::ggsave(
  filename = "../Results/Feature_processing/addEquivalent/count_id_Shorthouse_addEquivalent_KEGG.svg",
  plot = count_id_Shorthouse_addEquivalent_KEGG$Plot,
  width = 12,
  height = 8
)

count_id_Shorthouse_addEquivalent_CHEBI <- 
    MetaProViz::count_id(
        features_Shorthouse_addEquivalent, 
        delimiter = ";",
        "chebi",
        fill_colors = c(`No ID` = "#006384", `Single ID` = "#9F0162", `Multiple IDs` = "#BEBEBE"),
        path = "../Results/Feature_processing/addEquivalent"
    )
ggplot2::ggsave(
  filename = "../Results/Feature_processing/addEquivalent/count_id_Shorthouse_addEquivalent_CHEBI.svg",
  plot = count_id_Shorthouse_addEquivalent_CHEBI$Plot,
  width = 12,
  height = 8
)

count_id_Shorthouse_addEquivalent_PUBCHEM <- 
    MetaProViz::count_id(
        features_Shorthouse_addEquivalent, 
        delimiter = ";",
        "pubchem",
        fill_colors = c(`No ID` = "#006384", `Single ID` = "#9F0162", `Multiple IDs` = "#BEBEBE"),
        path = "../Results/Feature_processing/addEquivalent"
    )
ggplot2::ggsave(
  filename = "../Results/Feature_processing/addEquivalent/count_id_Shorthouse_addEquivalent_PUBCHEM.svg",
  plot = count_id_Shorthouse_addEquivalent_PUBCHEM$Plot,
  width = 12,
  height = 8
)
```

```{r}
# append the new "addEquivalent" ID counts

id_summary_Shorthouse <- append_id_counts(id_summary_Shorthouse, count_id_Shorthouse_addEquivalent_HMDB,    "HMDB_addEquivalent")
id_summary_Shorthouse <- append_id_counts(id_summary_Shorthouse, count_id_Shorthouse_addEquivalent_KEGG,    "KEGG_addEquivalent")
id_summary_Shorthouse <- append_id_counts(id_summary_Shorthouse, count_id_Shorthouse_addEquivalent_CHEBI,   "CHEBI_addEquivalent")
id_summary_Shorthouse <- append_id_counts(id_summary_Shorthouse, count_id_Shorthouse_addEquivalent_PUBCHEM, "PUBCHEM_addEquivalent")

## Calculate delta rows
delta_df_Shorthouse <- rbind(
  delta_df_Shorthouse,
  calculate_id_deltas(id_summary_Shorthouse, "original", "addEquivalent"),
  calculate_id_deltas(id_summary_Shorthouse, "cleaned", "addEquivalent"),
  calculate_id_deltas(id_summary_Shorthouse, "traverse", "addEquivalent")
)

## cleaned category dataframe
none_summary_Shorthouse <- append_none_summary(none_summary_Shorthouse, features_Shorthouse_addEquivalent, "addEquivalent")

## Calculate delta row
none_delta_Shorthouse <- rbind(
  none_delta_Shorthouse,
  calculate_none_delta(none_summary_Shorthouse, "original", "addEquivalent"),
  calculate_none_delta(none_summary_Shorthouse, "cleaned", "addEquivalent"),
  calculate_none_delta(none_summary_Shorthouse, "traverse", "addEquivalent")
)
```

```{r}
# save id summary
write.csv(
    id_summary_Shorthouse,
    "../Results/Feature_processing/id_summary/id_summary_Shorthouse.csv",
    row.names = TRUE,
    fileEncoding = "UTF-8"
)
write.csv(
    delta_df_Shorthouse,
    "../Results/Feature_processing/id_summary/delta_df_Shorthouse.csv",
    row.names = TRUE,
    fileEncoding = "UTF-8"
)
write.csv(
    none_summary_Shorthouse,
    "../Results/Feature_processing/id_summary/none_summary_Shorthouse.csv",
    row.names = TRUE,
    fileEncoding = "UTF-8"
)
write.csv(
    none_delta_Shorthouse,
    "../Results/Feature_processing/id_summary/none_delta_Shorthouse.csv",
    row.names = TRUE,
    fileEncoding = "UTF-8"
)
```

```{r}
# Plots
id_types <- c("HMDB", "KEGG", "PUBCHEM", "CHEBI")
id_type_colors <- c(
  HMDB = "#bbcc33",
  KEGG = "#99ddff",
  PUBCHEM = "#feaf16",
  CHEBI = "#d03293"
)

data_versions <- c("original", "cleaned", "traverse", "addEquivalent")

theme_no_grid <- theme_bw() +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )
mpv_theme <- theme_classic(base_size = 11) +
  theme(
    plot.title = element_text(size = 12, face = "bold"),
    axis.text.x = element_text(angle = 90, hjust = 1),
    legend.position = "right",
    legend.box = "vertical",
    legend.box.margin = margin(10, 10, 10, 10)
  )
plot_theme <- theme_no_grid + mpv_theme

id_summary_plot_df <- id_summary_Shorthouse %>%
  mutate(
    id_type = sub("_.*$", "", row_name),
    data_version = sub("^[^_]+_", "", row_name)
  ) %>%
  filter(id_type %in% id_types, data_version %in% data_versions) %>%
  mutate(
    id_type = factor(id_type, levels = id_types),
    data_version = factor(data_version, levels = data_versions)
  ) %>%
  complete(
    id_type,
    data_version,
    fill = list(no_ID = 0L, Single_ID = 0L, Multiple_IDs = 0L, Total_IDs = 0L)
  )

version_levels <- unlist(lapply(data_versions, function(v) paste(v, id_types, sep = " | ")))
type_levels <- unlist(lapply(id_types, function(t) paste(t, data_versions, sep = " | ")))

compact_bar_labels <- function(x) sub("^.*\\|\\s*", "", x)

version_group_labels <- tibble(
  data_version = data_versions,
  x = ((seq_along(data_versions) - 1) * length(id_types)) + (length(id_types) + 1) / 2
)

type_group_labels <- tibble(
  id_type = id_types,
  x = ((seq_along(id_types) - 1) * length(data_versions)) + (length(data_versions) + 1) / 2
)

version_group_annotation <- geom_text(
  data = version_group_labels,
  aes(x = x, y = -Inf, label = data_version),
  inherit.aes = FALSE,
  vjust = 3.0,
  fontface = "bold",
  size = 3.2
)

type_group_annotation <- geom_text(
  data = type_group_labels,
  aes(x = x, y = -Inf, label = id_type),
  inherit.aes = FALSE,
  vjust = 3.0,
  fontface = "bold",
  size = 3.2
)

bar_plot_theme <- plot_theme +
  theme(
    axis.text.x = element_text(angle = 0, hjust = 0.5),
    plot.margin = margin(5.5, 5.5, 28, 5.5)
  )


id_summary_plot_df <- id_summary_plot_df %>%
  mutate(
    group_by_version = factor(paste(as.character(data_version), as.character(id_type), sep = " | "), levels = version_levels),
    group_by_type = factor(paste(as.character(id_type), as.character(data_version), sep = " | "), levels = type_levels)
  )
```


```{r}
stacked_long_version <- id_summary_plot_df %>%
  pivot_longer(
    cols = c(no_ID, Single_ID, Multiple_IDs),
    names_to = "id_bucket",
    values_to = "count"
  ) %>%
  mutate(id_bucket = factor(id_bucket, levels = c("Multiple_IDs", "Single_ID", "no_ID"))) %>%
  group_by(group_by_version) %>%
  arrange(id_bucket, .by_group = TRUE) %>%
  mutate(label_y = cumsum(count) - count / 2) %>%
  ungroup()

stacked_long_type <- id_summary_plot_df %>%
  pivot_longer(
    cols = c(no_ID, Single_ID, Multiple_IDs),
    names_to = "id_bucket",
    values_to = "count"
  ) %>%
  mutate(id_bucket = factor(id_bucket, levels = c("Multiple_IDs", "Single_ID", "no_ID"))) %>%
  group_by(group_by_type) %>%
  arrange(id_bucket, .by_group = TRUE) %>%
  mutate(label_y = cumsum(count) - count / 2) %>%
  ungroup()

p_id_stacked_by_version <- ggplot(stacked_long_version, aes(x = group_by_version, y = count, fill = id_bucket)) +
  geom_col(width = 0.8) +
  geom_text(
    aes(y = label_y, label = ifelse(count == 0, "", count)),
    color = "black",
    size = 2.8
  ) +
  scale_x_discrete(drop = FALSE, labels = compact_bar_labels) +
  labs(
    title = "ID Summary by Data Version and ID Type",
    x = "Data Version and ID Type",
    y = "Feature Count",
    fill = "ID Count Class"
  ) +
  version_group_annotation +
  coord_cartesian(clip = "off") +
  bar_plot_theme

p_total_ids_by_version <- ggplot(id_summary_plot_df, aes(x = group_by_version, y = Total_IDs, fill = id_type)) +
  geom_col(width = 0.8, show.legend = FALSE) +
  geom_text(
    aes(y = pmax(Total_IDs - 5, 0), label = ifelse(Total_IDs == 0, "", Total_IDs)),
    color = "black",
    size = 2.8,
    fontface = "bold"
  ) +
  scale_fill_manual(values = id_type_colors) +
  scale_x_discrete(drop = FALSE, labels = compact_bar_labels) +
  labs(
    title = "Total IDs by Data Version and ID Type",
    x = "Data Version and ID Type",
    y = "Total_IDs"
  ) +
  version_group_annotation +
  coord_cartesian(clip = "off") +
  bar_plot_theme

p_id_stacked_by_type <- ggplot(stacked_long_type, aes(x = group_by_type, y = count, fill = id_bucket)) +
  geom_col(width = 0.8) +
  geom_text(
    aes(y = label_y, label = ifelse(count == 0, "", count)),
    color = "black",
    size = 2.8
  ) +
  scale_x_discrete(drop = FALSE, labels = compact_bar_labels) +
  labs(
    title = "ID Summary by ID Type and Data Version",
    x = "ID Type and Data Version",
    y = "Feature Count",
    fill = "ID Count Class"
  ) +
  type_group_annotation +
  coord_cartesian(clip = "off") +
  bar_plot_theme

p_total_ids_by_type <- ggplot(id_summary_plot_df, aes(x = group_by_type, y = Total_IDs, fill = id_type)) +
  geom_col(width = 0.8, show.legend = FALSE) +
  geom_text(
    aes(y = pmax(Total_IDs - 5, 0), label = ifelse(Total_IDs == 0, "", Total_IDs)),
    color = "black",
    size = 2.8,
    fontface = "bold"
  ) +
  scale_fill_manual(values = id_type_colors) +
  scale_x_discrete(drop = FALSE, labels = compact_bar_labels) +
  labs(
    title = "Total IDs by ID Type and Data Version",
    x = "ID Type and Data Version",
    y = "Total_IDs"
  ) +
  type_group_annotation +
  coord_cartesian(clip = "off") +
  bar_plot_theme

print(p_id_stacked_by_version)
print(p_total_ids_by_version)
print(p_id_stacked_by_type)
print(p_total_ids_by_type)
```
```{r}
ggsave(
  filename = "../Results/Feature_processing/id_summary_plots/id_summary_stacked_by_data_version.svg",
  plot = p_id_stacked_by_version,
  width = 16,
  height = 8
)
ggsave(
  filename = "../Results/Feature_processing/id_summary_plots/id_summary_total_ids_by_data_version.svg",
  plot = p_total_ids_by_version,
  width = 16,
  height = 8
)
ggsave(
  filename = "../Results/Feature_processing/id_summary_plots/id_summary_stacked_by_id_type.svg",
  plot = p_id_stacked_by_type,
  width = 16,
  height = 8
)
ggsave(
  filename = "../Results/Feature_processing/id_summary_plots/id_summary_total_ids_by_id_type.svg",
  plot = p_total_ids_by_type,
  width = 16,
  height = 8
)
```

```{r}
## ID summary plots with delta annotations (delta vs Original)

id_summary_with_delta_df <- id_summary_plot_df %>%
  group_by(id_type) %>%
  arrange(data_version, .by_group = TRUE) %>%
  mutate(
    delta_no_ID = no_ID - lag(no_ID, default = first(no_ID)),
    delta_Single_ID = Single_ID - lag(Single_ID, default = first(Single_ID)),
    delta_Multiple_IDs = Multiple_IDs - lag(Multiple_IDs, default = first(Multiple_IDs)),
    delta_Total_IDs = Total_IDs - lag(Total_IDs, default = first(Total_IDs))
  ) %>%
  ungroup()

stacked_long_version_delta <- id_summary_with_delta_df %>%
  pivot_longer(
    cols = c(no_ID, Single_ID, Multiple_IDs),
    names_to = "id_bucket",
    values_to = "count"
  ) %>%
  mutate(
    id_bucket = factor(id_bucket, levels = c("Multiple_IDs", "Single_ID", "no_ID")),
    delta_value = case_when(
      id_bucket == "no_ID" ~ delta_no_ID,
      id_bucket == "Single_ID" ~ delta_Single_ID,
      id_bucket == "Multiple_IDs" ~ delta_Multiple_IDs
    )
  ) %>%
  group_by(group_by_version) %>%
  arrange(id_bucket, .by_group = TRUE) %>%
  mutate(label_y = cumsum(count) - count / 2) %>%
  ungroup()

stacked_long_type_delta <- id_summary_with_delta_df %>%
  pivot_longer(
    cols = c(no_ID, Single_ID, Multiple_IDs),
    names_to = "id_bucket",
    values_to = "count"
  ) %>%
  mutate(
    id_bucket = factor(id_bucket, levels = c("Multiple_IDs", "Single_ID", "no_ID")),
    delta_value = case_when(
      id_bucket == "no_ID" ~ delta_no_ID,
      id_bucket == "Single_ID" ~ delta_Single_ID,
      id_bucket == "Multiple_IDs" ~ delta_Multiple_IDs
    )
  ) %>%
  group_by(group_by_type) %>%
  arrange(id_bucket, .by_group = TRUE) %>%
  mutate(label_y = cumsum(count) - count / 2) %>%
  ungroup()

p_id_stacked_by_version_with_delta <- ggplot(stacked_long_version_delta, aes(x = group_by_version, y = count, fill = id_bucket)) +
  geom_col(width = 0.8) +
  geom_text(
    aes(
      y = label_y,
      label = ifelse(count == 0 & delta_value == 0, "", paste0(count, "\nΔ", sprintf("%+d", delta_value)))
    ),
    color = "black",
    size = 2.5,
    lineheight = 0.9
  ) +
  scale_x_discrete(drop = FALSE, labels = compact_bar_labels) +
  labs(
    title = "ID Summary by Data Version and ID Type (with Δ vs Original)",
    x = "Data Version and ID Type",
    y = "Feature Count",
    fill = "ID Count Class"
  ) +
  version_group_annotation +
  coord_cartesian(clip = "off") +
  bar_plot_theme

p_total_ids_by_version_with_delta <- ggplot(id_summary_with_delta_df, aes(x = group_by_version, y = Total_IDs, fill = id_type)) +
  geom_col(width = 0.8, show.legend = FALSE) +
  geom_text(
    aes(
      y = pmax(Total_IDs - 8, 0),
      label = ifelse(Total_IDs == 0 & delta_Total_IDs == 0, "", paste0(Total_IDs, "\nΔ", sprintf("%+d", delta_Total_IDs)))
    ),
    color = "black",
    size = 2.6,
    fontface = "bold",
    lineheight = 0.9
  ) +
  scale_fill_manual(values = id_type_colors) +
  scale_x_discrete(drop = FALSE, labels = compact_bar_labels) +
  labs(
    title = "Total IDs by Data Version and ID Type (with Δ vs Original)",
    x = "Data Version and ID Type",
    y = "Total_IDs"
  ) +
  version_group_annotation +
  coord_cartesian(clip = "off") +
  bar_plot_theme

p_id_stacked_by_type_with_delta <- ggplot(stacked_long_type_delta, aes(x = group_by_type, y = count, fill = id_bucket)) +
  geom_col(width = 0.8) +
  geom_text(
    aes(
      y = label_y,
      label = ifelse(count == 0 & delta_value == 0, "", paste0(count, "\nΔ", sprintf("%+d", delta_value)))
    ),
    color = "black",
    size = 2.5,
    lineheight = 0.9
  ) +
  scale_x_discrete(drop = FALSE, labels = compact_bar_labels) +
  labs(
    title = "ID Summary by ID Type and Data Version (with Δ vs Original)",
    x = "ID Type and Data Version",
    y = "Feature Count",
    fill = "ID Count Class"
  ) +
  type_group_annotation +
  coord_cartesian(clip = "off") +
  bar_plot_theme

p_total_ids_by_type_with_delta <- ggplot(id_summary_with_delta_df, aes(x = group_by_type, y = Total_IDs, fill = id_type)) +
  geom_col(width = 0.8, show.legend = FALSE) +
  geom_text(
    aes(
      y = pmax(Total_IDs - 8, 0),
      label = ifelse(Total_IDs == 0 & delta_Total_IDs == 0, "", paste0(Total_IDs, "\nΔ", sprintf("%+d", delta_Total_IDs)))
    ),
    color = "black",
    size = 2.6,
    fontface = "bold",
    lineheight = 0.9
  ) +
  scale_fill_manual(values = id_type_colors) +
  scale_x_discrete(drop = FALSE, labels = compact_bar_labels) +
  labs(
    title = "Total IDs by ID Type and Data Version (with Δ vs Original)",
    x = "ID Type and Data Version",
    y = "Total_IDs"
  ) +
  type_group_annotation +
  coord_cartesian(clip = "off") +
  bar_plot_theme

print(p_id_stacked_by_version_with_delta)
print(p_total_ids_by_version_with_delta)
print(p_id_stacked_by_type_with_delta)
print(p_total_ids_by_type_with_delta)
```

```{r}
ggsave(
  filename = "../Results/Feature_processing/id_summary_plots/id_summary_stacked_by_data_version_with_delta.svg",
  plot = p_id_stacked_by_version_with_delta,
  width = 16,
  height = 8
)
ggsave(
  filename = "../Results/Feature_processing/id_summary_plots/id_summary_total_ids_by_data_version_with_delta.svg",
  plot = p_total_ids_by_version_with_delta,
  width = 16,
  height = 8
)
ggsave(
  filename = "../Results/Feature_processing/id_summary_plots/id_summary_stacked_by_id_type_with_delta.svg",
  plot = p_id_stacked_by_type_with_delta,
  width = 16,
  height = 8
)
ggsave(
  filename = "../Results/Feature_processing/id_summary_plots/id_summary_total_ids_by_id_type_with_delta.svg",
  plot = p_total_ids_by_type_with_delta,
  width = 16,
  height = 8
)

```

```{r}
baseline_steps <- tibble::tribble(
  ~from_version, ~to_version, ~step_label,
  "original", "cleaned", "original -> cleaned",
  "original", "traverse", "original -> traverse",
  "original", "addEquivalent", "original -> addEquivalent"
)

delta_plot_df <- delta_df_Shorthouse %>%
  mutate(parts = stringr::str_split(row_name, "_", simplify = TRUE)) %>%
  mutate(
    id_type = parts[, 2],
    from_version = parts[, 3],
    to_id_type = parts[, 4],
    to_version = parts[, 5]
  ) %>%
  select(-parts) %>%
  filter(
    id_type %in% id_types,
    from_version %in% data_versions,
    to_version %in% data_versions
  ) %>%
  left_join(baseline_steps, by = c("from_version", "to_version")) %>%
  filter(!is.na(step_label)) %>%
  mutate(
    id_type = factor(id_type, levels = id_types),
    step_label = factor(step_label, levels = baseline_steps$step_label)
  ) %>%
  mutate(across(c(no_ID, Single_ID, Multiple_IDs, Total_IDs), ~ tidyr::replace_na(., 0)))

## Plot 1: Total_IDs deltas vs Original (line plot)
delta_total_df <- delta_plot_df %>%
  filter(!is.na(Total_IDs))

p_delta_total_vs_original <- ggplot(delta_total_df, aes(x = step_label, y = Total_IDs, color = id_type, group = id_type)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
  geom_line(linewidth = 0.9) +
  geom_point(size = 2.4) +
  geom_text(
    aes(label = Total_IDs),
    vjust = ifelse(delta_total_df$Total_IDs >= 0, -0.55, 1.25),
    size = 2.8,
    color = "black"
  ) +
  labs(
    title = "Delta Total_IDs vs Original",
    x = "Comparison",
    y = "Delta Total_IDs",
    color = "ID Type"
  ) +
  plot_theme

## Plot 2: no_ID / Single_ID / Multiple_IDs deltas vs Original (line plot)
delta_comp_long <- delta_plot_df %>%
  select(id_type, step_label, no_ID, Single_ID, Multiple_IDs) %>%
  pivot_longer(
    cols = c(no_ID, Single_ID, Multiple_IDs),
    names_to = "metric",
    values_to = "delta"
  ) %>%
  filter(!is.na(delta)) %>%
  mutate(metric = factor(metric, levels = c("no_ID", "Single_ID", "Multiple_IDs")))

p_delta_comp_vs_original <- ggplot(delta_comp_long, aes(x = step_label, y = delta, color = metric, group = metric)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2) +
  geom_text(
    aes(label = ifelse(delta == 0, "", delta)),
    vjust = ifelse(delta_comp_long$delta >= 0, -0.45, 1.15),
    size = 2.3,
    color = "black"
  ) +
  facet_wrap(~ id_type, ncol = 2) +
  labs(
    title = "Delta Composition vs Original",
    x = "Comparison",
    y = "Delta Count",
    color = "Metric"
  ) +
  plot_theme

print(p_delta_total_vs_original)
print(p_delta_comp_vs_original)
```

```{r rebuild_feature_metadata}

features_Shorthouse_extended <- 
    read.csv(
        "../Results/Feature_processing/addEquivalent/features_Shorthouse_addEquivalent.csv", 
        row.names=1, 
        header = TRUE, 
        encoding="UTF-8", 
        check.names = FALSE
    )

Shorthouse_2022_extended <- Shorthouse_2022
Shorthouse_2022_extended$features <- features_Shorthouse_extended

save(Shorthouse_2022_extended, file = "../data/Shorthouse_2022_extended.RData")

```

```{r}
# clean up
gc()
rm(list = ls())
```

```{r}
sessionInfo()
```

